---
title: "Child Signals of Need in Utila, Honduras: An Exploratory Study"
execute: 
  echo: false
  warning: false
format: html
lightbox: true
editor: visual
tbl-cap-location: bottom
bibliography: signaling.bib
---

```{r setup, include=FALSE}
source("analysisforpaper.R", local = knitr::knit_global())

```

# Introduction

Despite the complexity of childhood and adolescence and their importance to adult outcomes, the behavioral strategies of children are often overlooked, especially when it comes to how they elicit support from parents and other caretakers [@gottlieb2000; @wells2003; @hirschfeld2002]. To gain support, children use signals to communicate their desires to parents. Infant signals of need, particularly infant cries, have been extensively studied in a range of populations [@barr1991; @fouts2005; @konner1974; @kruger2010; @kushnick2006; @meehan2017; @müller2021; @tronick1987], yet the full range of potential child signals of need, which likely extend beyond cries, has never been investigated in children of all ages. In particular, if and when children signal need with costly signals [@grafen1990; @spence1973; @zahavi1975a] has rarely been investigated [for one exception, see, @syme2023a].

In this paper, we use exploratory analysis on child signaling data from Utila Honduras to examine the hypothesis that children have evolved strategies to tailor their signals of need to 1) their valuation of a given resource, 2) the extent of conflicts of interest with caretakers, and 3) the presence of existing information about a child's need or potential to benefit from a resource so that they can credibly convey they are worth additional investment. Such strategies can lead to some combination of verbal requests for more support, facial sadness, crying, withdrawal, temper tantrums and self-harm, all of which may be effective in certain circumstances and are sometimes successful in eliciting more support from caretakers, despite a child's low status and formidability.

## Interdependence and investment within the family

Throughout human evolutionary history, individuals often found themselves in situations in which they received fitness-relevant benefits from others that could not easily be replaced [@apicella2012; @apicella2019; @hill2011; @tooby1996]. In these instances, natural selection is expected to have favored those who provided support to others if 1) such help meant one could expect to receive fitness benefits from those helped in the future and 2) the costs of providing help were outweighed by the benefits to the helper on average [@andré2010; @hamilton1964; @tooby1996; @trivers1971].

Care motivated by a high degree of fitness interdependence between offspring and at least one parent is seen in many animals [@clutton-brock1991; @webb1999]. However, humans are a uniquely interdependent species among the living primates [@tomasello2012] in that one’s fitness prospects depend greatly on a network of kin and other social allies [@apicella2019]. This need for others is most intense with children, who are born highly altricial and require many years of investment to develop to the point where they can obtain enough food to survive on their own and develop the social skills needed to succeed in their communities [@kaplan2009; @rosenberg2021; @schniter2015]. Much of this investment comes from parents [@kramer2018], However, other adults like grandparents, aunts, uncles, and non-relatives have also been found to provide significant investment in children in many societies [@kramer2018; @sear2008], something which was likely favored due to a combination of indirect fitness benefits, reciprocity, or reputation management [@gurven2000] + (Schniter Preprint (permission needed). Children also often serve as alloparents, helping to raise their younger siblings or provide for them through food preparation or collection [@hrdy2009; @meehan2009; @sear2008], despite still requiring substantial amounts of investment themselves [@kramer2018a].

## Conflicts of interest within the family

Although the fitness prospects of both children and familial caretakers are largely intertwined, evolutionary theory predicts that disagreements about the amount of investment from parents will be ubiquitous [@fouts2005; @godfray1995; @godfray1992; @trivers1974]. Most resources relevant to fitness outcomes are limited so that one unit of a resource invested in one area cannot be used in another. In response to this basic fact of life, humans, like other organisms, are expected to have evolved to efficiently allocate resources between somatic effort (i.e., growth and maintenance) and reproductive effort (i.e., mating and parenting effort) [@daly1983], something which often pits the fitness interests of parents against the fitness interests of offspring.

Conflicts of interest within families are largely structured by the likelihood an individual shares their genes with another. As a sexually reproducing species, humans have a 50% chance of passing down a given gene to each child. Each child, however, shares more genes through descent with themselves (100%) than with their full siblings (50% on average), half-siblings (25% on average), and stepsiblings (0%). Due to this asymmetry, children are often expected to attempt to elicit more investment from parents than is optimal from the perspective of the parent, while parents are expected to attempt to invest less than is optimal for each child in order to invest in their other children, whom parents are expected to value identically, all else being equal [@trivers1974].

For the same reason, siblings also often find themselves in conflicts of interest with one another, with each being expected to prefer more resources for themself than others. Sibling competition is evident across taxa and can range from relatively benign forms like strong begging competition in chicks [@caro2016] to siblicide in birds, mammals, and sharks [@salmon2014]. In humans, sibling competition is largely estimated by counts of siblings, and large numbers of siblings have been associated with decreased growth [@hadley2011; @hagen2006; @salmon2014], survival [@lawson2012; @meij2009; @strassmann2011], and reproductive success [@gibson2011; @mace1996] in multiple populations, highlighting the likely significance of sibling competition for resources, even if this does not always manifest with behavioral conflict [@bateson1994].

Potentially cut this paragraph: While parent-offspring conflict and sibling competition are likely ubiquitous, the resources being contested will vary across human groups. In relatively egalitarian and food constrained populations, conflict over resources is likely to be centered on food, which probably explains the negative impacts siblings have on child age-specific heights and weights among the Shuar of Ecuador [@hagen2009], although similar effects have also been found in industrialized populations [@lawson2008]. In societies with greater inequality, competition is more likely to be over wealth, as is the case with older brothers in pastoralist communities like the Gabbra and Arsi Oromo limiting the marriage prospects of laterborn sons [@gibson2011; @mace1996]. Compared to large-scale investments like bride price or inheritance, little is understood about smaller transfers of wealth to younger children despite the potential importance of such transfers to 1) facilitate the development of embodied capital outside of growth and fat reserves or 2) help children succeed in their peer groups.

Conflicts of interest between children and parents may also involve how much children invest in the household. As previously mentioned, each child is expected to prefer greater investment in oneself than in one’s siblings, while parents, who are equally related to all their children, should often be expected to favor a more even distribution [@trivers1974]. This can lead to conflict with older children who might prefer to spend their time in ways which are likely to result in future direct fitness benefits (e.g., finding romantic partners or forming strong bonds with friends) rather than taking care of their siblings or acquiring or preparing food [@lowe2003; @syme2023a]. Conflicts of interest over investment in the household also appear to extend to when a child starts having their own children [@kushnick2023; @emlen1995], with parents potentially benefiting from delaying the reproductive careers of certain offspring in order to gain more care for their other children [@flinn1989; @emlen1992] or speeding up a child’s departure from the household if the child’s presence limits how much caretakers can invest in other children they deem more worthy of investment [@kushnick2023].

## Signaling

In order to elicit more support, children often use signals. Signals are a key source of information for individuals in social species and are commonly used to indicate one's value as a mate, one’s formidability to competitors, or one’s worthiness of investment to parents and other social partners [@caro2016; @maynard-smith2004; @preuschoft2000; @setchell2005; @zahavi1975a]. Signals differ from cues in that signals were favored by natural selection due to their ability to alter the behavior or perceptions of others in ways that benefit the signaler, whereas cues are pieces of information organisms use to influence their behavior or physiological state that did not evolve for communicative effect [@maynardsmith1995]. Crying is likely a signal of need[^1], for instance, while bleeding would be a cue of need.

[^1]: We expect “signals of need” to actually be signals of resource valuation. However, we use “signal of need” here to distinguish signals that involve attempts at eliciting support through revealing private information about one’s vulnerabilities from the more commonly studied signals which reveal private information about one’s positive qualities (i.e., those related to partner choice and intrasexual competition).

When conflicts of interest over investment exist, there may be much to gain from deception (i.e., exaggerating or completely fabricating one’s quality or need) so that one can receive the benefits that come from being seen as having a particular quality without actually doing so [@zahavi1993]. Although the risks of being deceived can be substantial, signal receivers may still benefit from paying attention to signals so long as the information is credible enough to inform one’s behavior in ways that are adaptive on average [@bliegebird2001; @grafen1990]. These benefits coupled with the likelihood of strong selection against individuals who respond to fake signals results in the expectation that signals within species will often need to benefit both signaler and receiver to persist evolutionary [@hasson1994; @maynard-smith2004].

Credible signals[\[2\]](#_ftn2) can evolve and be maintained in multiple ways. Despite reasonable expectations that such circumstances will be rare [@zahavi1993], signals can be produced and trusted in instances when there are no conflicts of interest over what information is conveyed [@guilford1995; @maynardsmith1995]. They may also arise when signals cannot be faked (i.e., indexical signals) or in instances when there are greater net payoffs to signalers who have the quality being signaled to a greater extent than others [@maynardsmith1995]. Although the latter can arise through differential benefits [@penn2020; @tiokhin2016] or a combination of differential costs and benefits (Bergstrom preprint), more attention has been paid to the importance of differential costs alone, with costly signaling being a possible pathway to credible signaling whenever signals are prohibitively costly for those who might fake the signal but still worth making for those with the underlying trait [@maynard-smith2004].

Credible signals work by revealing private information about one’s qualities to others. In the context of child signals of need, a child has access to information about their internal state, which is unavailable to others, that allows them to estimate their valuation of additional investment. By signaling in a credible way, the child reveals their resource valuation to potential caretakers who may be skeptical of the child’s claims due to the common conflicts of interest described above. With the information revealed by the signal, caretakers should generally be more likely to provide additional support due to their fitness interest in the child’s wellbeing. When this is the case, all parties are better off than they would be without the signal, with caretakers benefitting from the revelation of fitness-relevant information and the child benefitting from the increased likelihood of a beneficial response as long as the benefits outweigh the costs of the signal to child and parent.

However, even if a child was completely successful in conveying their needs to caretakers, it may not be enough to result in a beneficial response. This is because belief in one’s child’s need is not the only factor determining the form and amount of support given by parents. Instead, parents are expected to weigh the costs and benefits of support with other investment options [@daly1987; @daly1983]. Ancestrally, this likely often meant evaluating the viability of the child as well as how much a particular resource benefited one child relative to other children or future children [@clutton-brock1991; @daly1983; @dawkins1976; @hagen1999]. The former is thought to have resulted in children having to demonstrate their vigor early in life [@soltis2004], while the latter is expected to result in context-specific support [@beaulieu2008; @frankenhuis2022]. For example, one unit of a resource is often expected to be worth more to a newborn than an adolescent, favoring that investment flowing to the newborn all else being equal. However, if resources are in short supply, parents may benefit more from investing more in older children as they are closer to reaching adulthood and have higher reproductive value [@daly1983]. Additionally, a parent may also decide against investing in a child despite deeming investment in a child worthwhile if time is not right (e.g., if a mother preferred to breastfeed during rest rather than foraging) [@altmann1980]. When this is the case, outward conflicts may exaggerate the extent of a conflict of interest and involve parents teaching children ways to behave that benefit them both [@altmann1980].

### Crying

Crying is the most studied way by which children may signal need. Crying is a human universal [@gracanin2018] with few suspected non-signaling functions outside of the role vocal cries may play in the development of vocalization [@wells2003] and the functions tears have in terms of cleaning, protecting, and lubricating the eye [@murube2009; @sznycer]. Hunger, pain, anger, fear, and separation are all common causes of crying [@vingerhoets2013; @wells2003], although tear production also often happens with joyful events [@sznycer; @vingerhoets2013; @zickfeld].

Infant cries have been found to result in increased amounts of care, resources, and attention they receive and increased proximity to parents [@lin2012]. For example, Karo Batak infants who cried more were breastfed more, and the length of crying bouts predicted the time mothers breastfed [@kushnick2006]. Similarly, !Kung San mothers frequently respond to crying infants by providing comfort and breastfeeding access, with comforting behaviors, in general, being far more common than one’s meant to restrict infant behavior [@kruger2010]. Although this study found mothers were the most common responders, siblings, fathers, and other adults frequently stepped in so that 88% of crying bouts were responded to, with most of the cries not responded to lasting under ten seconds. Similar responsiveness has also been reported among the Aka [@hewlett1998], Mbendjele hunter-gatherers [@chaudhary2024], the Efe [@tronick1987], and the Ache, among whom young children have been described as “win\[ning\] most conflicts with parents simply by crying and whining” [@kaplan1987]. Studies examining the potential for long-term benefits are rare and difficult to achieve; however, during a time of high infant mortality in a sample of 13 Massai infants, those with easy temperaments were more likely to die (5/7) than those with difficult temperaments (1/6), something which is consistent with signals resulting in beneficial treatment [@devries1987].

Crying does not always succeed in eliciting benefits, however. Compared to the previously mentioned groups, the Ngandu [@hewlett1998] and the Hadza [@blurtonjones1993] have been found to be more likely to ignore the cries of infants and young children, and concerns over the risk of spoiling children by being too responsive seem to be reasonably common in some Western populations [@barnett2010; @burchinal2010; @chaudhary2024; @frankenhuis2022; @maute2018]. Crying can also lead to negative responses beyond being ignored [@barr1990; @soltis2004; @zeifman2017]. It has been found to increase parental distress, the likelihood of maternal depression, and the child’s risk of maltreatment [@frodi1985; @fujiwara2011; @howard2006; @lin2012; @soltis2004]. For example, in a sample of 3,259 parents of Dutch infants under 6 months old, 5.6% self-reported smothering, slapping, or shaking their babies in response to their crying, outcomes which were more likely among parents who were single, unemployed, or deemed their infant’s crying to be excessive [@reijneveld2004]. Excessive infant crying, i.e., colic, has also been found to be a risk factor for multiple negative outcomes [@zeevenhooven2018], including abuse, shaken baby syndrome [@kurth2011], and earlier cessation of breastfeeding [@howard2006].

Although widely accepted as a signal, there is no consensus on why emotional crying may work to elicit positive responses. One possibility is that the physiological costs of crying for newborns and infants [@rao1997; @thureen1998] may be prohibitively costly for faking to pay off. When this is the case, crying may serve as a credible of vigor [@furlow1997; @lummaa1998; @soltis2004; @wells2003], something which may be particularly important early in life due to increased risk of infanticide for children who have has poor survival prospects, due to either health conditions or resource constraints [@daly1984; @soltis2004].

Potentially combine and condense with paragraph below: As children age, their reproductive value increases due to reduced odds of mortality, leading to a decreased risk of infanticide (Hagen 2004 comment). For this reason, there is likely a transition period in which crying is primarily meant to provide information about how much a child values further investment. Many models have been put forward which attempt to explain human crying and begging in other species as credible signals of need due to the differential costs and benefits of crying and analogous vocalizations in other species as deterring the exaggeration of need [@godfray1991; @kilner1997; @parker2002]. Although differing in their details, a common theme is that the indirect fitness costs signalers incur from imposing costs on that parents and siblings through begging (be them energetic costs, survival costs, or reduced likelihood of future siblings) creates a floor to possible deception but falls short of preventing it [@daly1990; @wells2003].

For older children, the relatively insignificant energetic costs of crying may not be enough to result in credible signals in times of substantial conflicts of interest. For this reason, crying among older children may be more similar to adult crying, something which has been found to result in increased support and empathy but sometimes come at the cost of reputation and the potential for stigmatization [@cornelius2001; @cretser1982; @hendriks2006; @jesser1989; @vingerhoets2016]. Such costs may be a means by which crying can still be a credible signal [@sznycer] with moderate conflicts of interest if they dissuade potential fakers from making the signal.

### Withdrawal

Withdrawal has been suggested to be an underappreciated way those with low social power may credibly signal need [@syme2023a]. Due to their lower status, children have limited ways to change how others treat them. However, as children 1) are a major component of parents' inclusive fitness and 2) can help family members through allocare and other forms of labor, they provide benefits that are valuable and hard to replace. By withholding these benefits or putting the future of these benefits into question, a child simultaneously delays the benefits they get from this cooperation while imposing costs on those with a fitness interest in the child. As these costs are likely lower for those truly in need than those who are not, withdrawal may be a credible way to signal need if the costs to faking the signal make it unprofitable. In these cases, the end result is analogous to a labor strike, in which the signaler conveys to others their situation demands more support, and if they want to continue receiving benefits from the child, they need to upregulate their support of the child.

One of the key proximate mechanisms behind withdrawal is sadness or low mood [@scherer1994]\[\[3\]\](#\_ftn3). Sadness is a type of negative affect that is generally thought to arise from loss [@horwitz2017; @levenson1999; @smith1993], especially when one has little power to reverse or mitigate the loss on one’s own [@emanuel2023; @gunderson2021]. Sadness and low mood, like other forms of psychological pain, probably evolved to motivate victims of adversity to shift their attention to the causes of adversity to mitigate its negative fitness consequences and to learn to solve or avoid future such adverse events [@andrews2009; @nesse1990; @thornhill1989]\[\[MG12\]\](#\_msocom_12). Low mood, for example, is often thought to be adaptive in situations when activity is dangerous or unlikely to result in benefits, with anhedonia, pessimism, and fatigue all helping reduce behavior that may not be worth the costs or that may cause additional problems [@nesse2019; @nettle2009]. It may also encourage the reappraisal of one’s circumstances after new information comes to light that one’s current strategy is not working [@nesse2009].

Through the reduction in behavior and attention in one’s usual interests that can accompany sadness and low-mood [@nettle2009], one’s relatives and other social partners would likely gain some indication, in the form of cues, that the sad individual was experiencing adversity and could benefit from increased support. In addition to this cue-based information, sadness also results in facial expressions and crying, both of which probably evolved to indicate need [@balsters2013; @bowlby1980; @reed2017; @darwin1872]. Such expressions seem to be important to conveying this information as is seen by responses to those who cannot produce certain elements of facial expressions like tear production [@hasson2009; @nesse2019] or certain facial movements [@bogart2012; @bogart2010]. However, facial expressions, along with tears, may not be enough to credibly signal need in times of significant conflicts of interest. In these instances, depression and suicidal behavior have been suggested to function as ways to credibly signal need due to the expectation that such states are costlier for those facing adversity than those who are not (for more detail on these approaches see: [@hagen2003; @rosenthal1993; @syme2016; @watson2002; @gaffney2022].

### Temper tantrums and self-harm

Temper tantrums may also be adaptive strategies by which children can elicit more support from parents. They are common in children from 18 months to 4 years old [@potegal2003], and similar behaviors have also been reported in other young animals, often surrounding feeding or weaning conflict [@altmann1980; @goodall1986; @schaller1964]. Temper tantrums vary in their severity, ranging from crying, shrieking, and screaming to violent bodily movements, aggression, and self-harm [@potegal2003]. They have been reported in many societies with different subsistence systems, political organization, and norms surrounding anger and parenting [@albino1956; @banisalameh2021; @fouts2005; @henderson1966; @hilger1957; @konner1974; @ward1970], suggesting they may be a human universal. As with the other proposed signals of need, they are capable of eliciting a wide range of responses, including pro-child responses, neutral responses (e.g., parents ignoring the child) [@hilger1957; @konner1974], and negative responses like abuse [@schmitt1987].

As with the vocalizations that come with crying, temper tantrums might motivate caretakers to upregulate their support of the signaler due to the energetic costs of tantrums or the increased risk of predation stemming from the noisy display, both of which might result in credible signals if they are costly enough to not be worth faking [@lummaa1998]. In addition, tantrums also impose direct costs on caretakers who may be less able to do fitness enhancing tasks [@altmann1980] like finding or preparing food or taking care of other children.

Another possibility is that temper tantrums carry the risk or implication of offspring self-harm. For example, Utilian mothers reported in focus groups that some temper tantrums involved children banging their heads on things when they do not get what they want, something that has been reported in other populations [@belden2008; @burrows1953; @daniels2012; @dennis1940; @vandenakker2022]. Similarly, young chimpanzees (*Pan troglodytes*) have been observed to beat their heads on the ground when refused access to breast milk [@goodall1986], and American white pelican chicks (*Pelecanus erythrorhynchos*) have been found to throw themselves on the ground, flail violently, and bite at their own wings when begging for food [@schaller1964]. Such behaviors have been suggested to follow similar logic to depression and suicide attempts as they are likely relatively less costly to those who are truly in dire need compared to those who are not to the point faking is not worth the costs [@hagen2008]. In other words, the risk of injury from violent temper tantrums carries greater fitness costs to a child who is receiving enough support to develop high levels of embodied capital than it does to a child who is neglected to the point where their survival prospects are low. In this scenario, the risk of harm is likely not worth the potential benefits of increased support to the child receiving substantial investment, but it may be for the neglected child who may have few other options to compel others to invest in them.

## Study Aims

This study was designed to help establish the natural history of child signaling behavior in Utila, Honduras, by quantifying the frequencies of common child signals of need and various types of parental responses. From this, we sought to use signaling theory to examine 1) the causes of child signals of need, 2) how and why signal costs might vary, 3) how parents respond to indications of child need, 4) how signaling behavior changes with age and sex, 5) and if there were differences in child signaling and parental responses two groups who face different levels of social and economic adversity.

A major challenge when studying signals of need is the fact that such signaling is expected to 1) be more common in times when individuals can benefit from increased investment (e.g., when they are facing major adversity or when a fitness-relevant opportunity arises) and 2) result in benefits to the signaler on average. This hypothisized causal, bi-directional relationship between investment and signaling makes it difficult to isolate the effects of signaling behavior on investment in children when using standard measures of child welfare (i.e., BMI, health metrics, and height for age). When experiments are not appropriate, longitudinal studies with carefully selected variables can help address this [@arjas2004]. However, our small sample size with signaling data and anthropometric data at two time periods limits our ability to do this in this study. For this reason, all analyses for this study were exploratory and meant to inform future studies of child signaling.

# Methods

## Study population

This study was done in Utila, Honduras, one of the country’s three major Bay Islands in the Western Caribbean. Participants in the study largely fit into two groups who face different levels of economic and social adversity, 1) a long-standing English-speaking community primarily made up of the descendants of expats and 2) a relatively new community of Spanish-speaking immigrants from the mainland. Compared to those born on Utila, immigrants from the mainland [@davis2014] are more likely to live in crowded and unsanitary conditions, face active discrimination [@garcia2017], and experience economic uncertainty due the predominance of low paying, seasonal jobs primarily related to the island’s diving tourism industry [@garcia2018; @stonich2000]. Such inequality appears to be increasing in response to recent growth in the tourism industry, with the majority of wealth from tourism being enjoyed by a small portion of expat islanders and westerners [@davis2014].

During focus groups held in July 2022, parents indicated high levels of sibling competition and various ways children might signal need. Both Spanish-speaking and English-speaking moms reported that fighting between siblings was extremely common, with fights over toys, gifts, and phone time being the most reported causes. In both populations, crying and various forms of withdrawal were listed as the most common responses to a child not receiving investment or support, followed by temper tantrums and making messes in their home. Spanish-speaking mothers reported that refusing to do homework was a common tactic, while English-speaking mothers emphasized running away and self-harm (e.g., a child slamming their head on the ground) to a greater degree. In response to these hypothesized signals, discipline was common. Although the mothers generally reported that children were treated equally, needier children were described as getting more resources from parents, something that may be partially mediated by differences in signaling among children in addition to information caretakers gained from cues.

## Sampling

Participants for this study were recruited by a variety of means. JH recruited early participants through convenience and snowball sampling. Participants were also recruited via Facebook posts and fliers which were sent home with children the Utila's Methodist school.

In all forms of recruitment, adult participants were informed of the purpose of the study, what they would be asked to do, and the inclusion criteria, which included being over 18, residing on Utila, and having a child between 6 and 18 years old.

XXXXXXXXCHIld RecruitmentXXXXXXXXXXX

## Participants

All participants for the survey portion of the study met the inclusion criteria. Most participants were mothers of at least one biological child in the study, with the participant being the biological mother of the child in xxx out of xxxx caregiver/child dyads.

Child participants for the anthropometric portion of the study were at least 5 years of age. In the majority of instances, child participants had a primary caretaker who took the survey portion of the study first, although some children participated before parents. To have their data included in this study, a caretaker in their household must have taken the survey.

## Ethics

All adult participants provided informed consent. Surveys were designed to last about an hour plus 20 minutes for each additional child. Adults were paid \$XXXXX for the survey portion of the study.

xxxxxxxxxxxxxChild consent infoxxxxxxxxxxxxxx.

This study was approved by the xxxxxxxIRBINFOxxxxxxxxxxxxx.

## Survey

The survey data for this study come from a larger project run by the Utila Child Health Project. Data was collected by either Jess Hlay, Izabel Rodriguez, Daniela Mejia, Angela Garcia, or Caroline Smith who read the survey questions to participants and recorded their answers. Participants could choose to do the survey in English or Spanish.

### Frequency of child signaling, alloparenting contributions, and conflict with caretaker

We used 6-point Likert scales to get caregiver reports on the frequencies of 1) various types of child signaling, 2) interpersonal conflict with each participating child, 3) and child alloparenting behavior. See @tbl-freqvars for exact wording. Caretakers were often not asked about child alloparenting behavior if a child was an only child. For this reason, we gave children an alloparenting frequency of never if they were the only child within a household.

|                                 |                                                                                                                   |
|--------------------------|----------------------------------------------|
| Variable                        | Survey question wording                                                                                           |
| Crying Frequency                | How often does your child cry?                                                                                    |
| Sad Frequency                   | How often is your child sad?                                                                                      |
| Tantrum Frequency               | How often does your child throw temper tantrums? (e.g., scream, throws things, or throws themselves on the floor) |
| Running Away Frequency          | How often does your child leave the house without telling you when they get mad?                                  |
| Parent/Child Conflict Frequency | How often do you have conflicts with your child?                                                                  |
| Child Alloparenting Frequency   | How often does your child take care of their other siblings?                                                      |

: Variable names and question wording for measures about the frequency of behaviors. All questions had the following labels attached to each point on its Likert scale: Never; Once a month or less; More than once a month but less than once a week; More than once a week but not daily; Daily; Multiple times per day. {#tbl-freqvars}

As these responses reflect frequencies of behavior, we converted these to counts which reflected the number of times per month a child engaged in these behaviors for analyses. The original wording of the Likert scale and the frequency conversion are shown in @tbl-convertedfreqs.

|                                                  |                               |
|------------------------------------|------------------------------------|
| Likert scale wording                             | Count value (times per month) |
| Never                                            | 0                             |
| Once a month or less                             | 1                             |
| More than once a month but less than once a week | 3                             |
| More than once a week but not daily              | 8                             |
| Daily                                            | 30                            |
| Multiple times per day                           | 60                            |

: The conversions from Likert scale to count value for the frequencies of child behaviors included in this study. Variables created with this scheme include the frequencies of sadness, crying, temper tantrums, conflict with primary caretakers, and child alloparenting effort.

From these count variables, we created three additional signaling variables to be used, primarily, as our main dependent variables in analyses. The frequency of signaling was the sum of the frequencies of child sadness, crying, and temper tantrums. Maximum signal frequency was the frequency of the most common of these three signals. Signal cost was the sum of the frequencies of these three signals weighted by their relative costliness (i.e., Signal cost = frequency of sadness + frequency of crying \* 2 + frequency of temper tantrums \* 3). Although we collected data on the frequency of running away, it was not included in the creation of any of these variables due to uncertainty over whether it was likely to be a credible signal due to the relative safety of Utila compared to other areas in which running away has been found to be a successful strategy for children.

We also calculated the summed frequency of child alloparenting contributions by other children within the household.

### Relative neediness of children and relative investment in children within the family

We also used 6-point Likert scales to get parental reports of 1) the relative neediness of each participating child and 2) the relative investment they provide to each participating child compared to the rest of their children. See Table @tbl-relvars for exact wording.

|                     |                                                                                                  |
|--------------------------|----------------------------------------------|
| Variable            | Survey question wording                                                                          |
| Relative Need       | How much time, effort, and resources does this child need compared to your other children?       |
| Relative Investment | How much time, effort, and resources do you put into this child compared to your other children? |

: Variable names and question wording for measures about relative need and investment. Both questions had the following labels attached to each point on its Likert scale: Much less; Less; The same amount; More; Much more. {#tbl-relvars}

### Causes and responses to the last instance of child signaling

We asked caretakers about the last time each participating child was sad or cried. This involved asking what the caretaker thought caused the emotion or display and recording their answer in a text box. We then asked how the parent responded to their child. The researcher giving the survey would then bin responses into at least one of 31 categories which included positive, negative, and neutral responses (from the child's perspective). For the full list of options see Table S1.

From this data, we created a parental response variable, which was coded as 1 for a positive response, 0 for a neutral response, and -1 for a negative response (all from the child's perspective). MG also coded each case for whether or not the cause had involved: 1) the child being asked to do an unwanted task, 2) punishment, 3) conflict within the family, and 4) discomfort, pain, injury, or illness.

### Perceptions of child susceptibility to illness (or current susceptibility to illness)

Caretakers were also asked three questions using a five point Likert scale about their child's susceptibility to illness. We inverted the scores on two measures so that a score of 5 corresponded with the lowest susceptibility to illness before we averaged the scores of these to create a mean illness susceptibility variable. In the `r sum((!is.na(modeldf$IllnessSusceptibility2)) & is.na(modeldf$IllnessSusceptibility3))` instance in which we only data for 2 of the measures, we still calculated the mean but used a numerator of 2. See Table @tbl-healthvars for the wording of these variables.

|                        |                                                                                       |
|--------------------------|----------------------------------------------|
| Variable               | Survey question wording                                                               |
| IllnessSusceptibility1 | If an illness is 'going around', my child will get it.                                |
| IllnessSusceptibility2 | In general, my child is very susceptible to colds, flu, and other infectious diseases |
| IllnessSusceptibility3 | My child’s immune system protects him/her from most illnesses that other kids get.    |

: Variable names and question wording for measures about a child's current likelihood of becoming sick. All three questions were asked using a Likert scale from 1-5 where 1 was labeled "Strongly disagree" and 5 was labeled "Strongly agree." The scores for IllnessSusceptibility1 and IllnessSusceptibility2 were inverted so that 5 corresponded with the lowest susceptibility to illness. {#tbl-healthvars}

### Demographic questions

The survey also included many demographic questions about the survey taker and their households. Those used for this study included the (1) ages of those in the household, (2) sexes of those in the household, (3) partnership status of the primary caregiver, (4) number of adults who provide childcare within the household, (5) highest level of education of the primary caregiver, (5) immigration status (in regards to Utila), and (6) monthly household income. This data allowed us to calculate the number of children (total, younger, and older) and adults within the household.

## Child anthropometrics

xxxxxxxxxxDescribe camps brieflyxxxxxxxxxxxxxx - include dates and how we handled those. For cross-sectional analyses, we used the 2023 anthropometric values due to them being closer in time to the caretaker surveys.

All anthropometric measurements were collected in accordance with anthropometric standards [@lohman1988] and involved taking the mean of two measurements. Before being measured, participants were asked to remove their shoes and heavy articles of clothing. Height was measured to the nearest millimeter using an anthropometer, and weight was measured to the nearest 100 g with a digital scale (model name) that used within a building with a concrete foundation. All participants were wearing lightweight clothing, so we made no adjustments for clothing weight. Grip strength was measured using a dynamometer (model name), with participants seated/standing? and told to xxxxxxxx. XXXX time was passed between the first and second measurements.

Triceps and subscapular skinfold thicknesses were measured using skinfold calipers. Measurements were taken directly on the skin. The triceps measurement was taken at the midpoint of the upper arm. The subscapular measurement occurred immediately below the inferior angle of the scapula, being picked up diagonally at the natural fold line of the skin [@antóna]. Flexed bicep circumference was measured perpendicular to the long axis of the upper arm. For this measure participants were asked to sit/stand? an position their upper arm perpendicular to their body with their elbow being help at a 90° angle, before flexing their clenched fist towards themselves as much as they could.

We computed each child's height, weight, and BMI Zscores with respect to the CDC child growth standards [@kuczmarski2002].

For our independent variables of skinfold thickness, flexed bicep circumference, and grip strength on signaling behavior, we used the residuals from generalized additive models which modeled the effect of sex and age on each measurement using the mgcv package [@wood2011]. The resulting variables which represent how each child differed from the average child for their given sex and age. We also centered and scaled the skinfold thickness residuals before summing them to create an overall body fat variable.

## Statistical analyses

All statistical analyses were exploratory and involved fitting many models with different combinations of independent and dependent variables. While were cognizant of the risk of overfitting the data, it is still more likely than if the study was entirely confirmatory [@tukey1977]. The payoff to this approach is the identification of theoretically relevant effects that should be tested in the future.

We fit generalized linear mixed models for exploratory analyses with signal frequencies and summed signal cost as dependent variables. As our signaling variables involved over-dispersed count data, we used negative binomial regression. This was done in R with the GLMM package [@brooks2017], with household as a random effect. Control variables differed in some models, but child age and sex were always included. Specific control variables are mentioned in the results.

For the models examining 1) parental responses to the last reported time each participating child was sad or cried, 2) the relative neediness of the child, and 3) the investment a child received compared to the other children in the household, we fit ordinal logistic regression models using two methods. The first used the polr function in the MASS package [@venables2002]. The results were then interpreted with clustered standard errors for household [@zeileis2002]. Due to the risk of overfitting that comes with having too few clusters [@cameron2015], we also fit the same models using BRMS [@bürkner2017] with household as a random effect.

## Data availability

The anonymized data used in the study are available at xxxxxxxLINKxxxxxxx. Study code is available at xxxxxxLINKxxxxxxxxxx.

# Results

`r dplyr::n_distinct(modeldf$householdID)` primary caretakers took the survey and provided signaling data for at least one child. `r nrow(modeldf)` children had frequency data for at least 1 signal, with `r nrow(modeldf_FULLSIG)` having data for all 3. Among those with at least one record of signaling frequency, `r sex_table[1]` were female and `r sex_table[2]` were male. Anthropometric and signaling data exist for `r nrow(modeldf_FULLANTHROPOMETRICS)` children, of whom `r sex_table2[1]` were female and `r sex_table2[2]` were male. For summary statistics of child-level variables see @tbl-sumtab_children. For summary statistics of caretaker and household-level variables see @tbl-sumtab_adults.

```{r sumtab_children}
#| label: tbl-sumtab_children
#| tbl-cap: Summary statistics for child study variables.
vars_c <- c(
  ChildAge = "Age (years)",
  SadFreqN = "Sad frequency (times per month)",
  CryFreqN = "Crying frequency (times per month)",
  TantrumFreqN = "Temper tantrum frequency (times per month)",
  AlloparentingFreqN = "Frequency of child's own alloparetning effort (times per month)",
  ConflictFreqN = "Frequency of conflict with focal caretaker (times per month)",
  SignalFreq = "Frequency of signals summed (times per month)",
  SignalCost = "Signal cost",
  SignalFreqMax = "Frequency of most common signal (times per month)",
  # IllnessSusceptibility1 = "Illness Susceptibility1",
  # IllnessSusceptibility2 = "Illness Susceptibility2",
  # IllnessSusceptibility3 = "Illness Susceptibility3",
  IllnessSusceptibilityMean = "Illness avoidance index",
  HeightZ = "Height Z-score (CDC)",
  WeightZ = "Weight Z-score (CDC)",
  BMIZ = "BMI Z-score (CDC)",
  SubscapMean = "Subscapular skinfold thickness (mm)",
  SubscapR = "Subscapular skinfold residuals",
  TricepMean = "Tricep skinfold thickness (mm)",
  TricepR = "Tricep skinfold residuals",
  BodyFat =  "Body fat composite"
)

kable(custom.summarize(modeldf, vars_c, facvar = "Sex")) %>% kable_styling()

#kable(custom.summarize(d2, vars), caption = "Summary statistics for child study variables.") %>% kable_styling()

```

```{r sumtab_adults}
#| label: tbl-sumtab_adults
#| tbl-cap: Summary statistics for adult and household study variables.

vars_c <- c(
  CaregiverAge = "Age (years)",
  EducationLevelYears = "Education level (years)",
  IncomeCategoryN = "Monthly household income (Lempira)",
  number_adults = "Adults in household",
  AdultsChildcare = "Adults who provide childcare"
)

kable(custom.summarize(caregivers, vars_c)) %>% kable_styling()

#kable(custom.summarize(d2, vars), caption = "Summary statistics for child study variables.") %>% kable_styling()

```

## Distributions of signaling behavior, alloparenting effort, and conflict

Child sadness, crying, and temper tantrums were common. Caregivers reported that `r sum(na.omit(modeldf$SadFreqN >= 8))` (`r signif(sum((na.omit(modeldf$SadFreqN >= 8)) / sum(!is.na(modeldf$SadFreqN)) * 100), 2)`%) children were sad at least once a week, `r sum(na.omit(modeldf$CryFreqN >= 8))` (`r signif(sum((na.omit(modeldf$CryFreqN >= 8)) / sum(!is.na(modeldf$CryFreqN)) * 100), 2)`%) cried at least more than once a week, and `r sum(na.omit(modeldf$TantrumFreqN >= 8))` (`r signif(sum((na.omit(modeldf$TantrumFreqN >= 8)) / sum(!is.na(modeldf$TantrumFreqN)) * 100), 2)`%) threw temper tantrums at least once a week. A small subset of children engaged in these signals/states at least daily (sadness = `r sum(na.omit(modeldf$SadFreqN >= 30))` children \[`r signif(sum((na.omit(modeldf$SadFreqN >= 30)) / sum(!is.na(modeldf$SadFreqN)) * 100), 2)`%\], crying = `r sum(na.omit(modeldf$CryFreqN >= 30))` children \[`r signif(sum((na.omit(modeldf$CryFreqN >= 30)) / sum(!is.na(modeldf$CryFreqN)) * 100), 2)`%\], and tantrums = `r sum(na.omit(modeldf$TantrumFreqN >= 30))` children \[`r signif(sum((na.omit(modeldf$TantrumFreqN >= 30)) / sum(!is.na(modeldf$TantrumFreqN)) * 100), 2)`%\]). However, many children were reported to never engage in these signals/states (sadness = `r sum(na.omit(modeldf$SadFreqN == 0))` children \[`r signif(sum((na.omit(modeldf$SadFreqN == 00)) / sum(!is.na(modeldf$SadFreqN)) * 100), 2)`%\], crying = `r sum(na.omit(modeldf$CryFreqN == 0))` children \[`r signif(sum((na.omit(modeldf$CryFreqN == 0)) / sum(!is.na(modeldf$CryFreqN)) * 100), 2)`%\], and tantrums = `r sum(na.omit(modeldf$TantrumFreqN == 0))` children \[`r signif(sum((na.omit(modeldf$TantrumFreqN == 0)) / sum(!is.na(modeldf$TantrumFreqN)) * 100), 2)`%\]). See @fig-barplot_SignalFreq for plot showing signaling frequencies and @fig-signal_alluvial_plot for plot showing the number of individuals who engaged in specific combinations of signal frequencies..

Conflict between children and caretakers was also common, with `r sum(na.omit(modeldf$ConflictFreqN >= 8))` children (`r signif(sum((na.omit(modeldf$ConflictFreqN >= 8)) / sum(!is.na(modeldf$ConflictFreqN)) * 100), 2)`%) having conflicts with parents more than once a week. As with the frequencies of various types of signaling, many children (`r sum(na.omit(modeldf$ConflictFreqN == 0))` \[`r signif(sum((na.omit(modeldf$ConflictFreqN == 0)) / sum(!is.na(modeldf$ConflictFreqN)) * 100), 2)`%\]) were described as never having conflicts with parents. See @fig-barplot_ConflictFreq for plot showing the frequencies of conflict between caretakers and children.

The frequency of alloparenting effort was bimodally distributed, with `r sum(na.omit(modeldf$AlloparentingFreqN == 0))` children (`r signif(sum((na.omit(modeldf$AlloparentingFreqN == 0)) / sum(!is.na(modeldf$AlloparentingFreqN)) * 100), 2)`%) being reported to never engage in it (`r sum(na.omit(modeldf$NumberOfChildren==1))` instances were due to a child being an only child) and `r sum(na.omit(modeldf$AlloparentingFreqN >= 30))` children (`r signif(sum((na.omit(modeldf$AlloparentingFreqN >= 30)) / sum(!is.na(modeldf$AlloparentingFreqN)) * 100), 2)`%) being reported to help with childcare at least once a day. See @fig-barplot_AlloparentingFreq for plot showing the frequencies child alloparenting effort.

::: {#fig-barplot_SignalFreq}
```{r}
barplot_SignalFreq
```

Stacked bar plot showing the frequencies of child sadness, crying, and temper tantrums per month as reported by primary caretaker. Frequency key: 0 corresponds to "never," \<=1 corresponds to "once a month or less," 2-3 corresponds to "more than once a month but less than once a week," 4-29 corresponds to "more than once a week but not daily," 30 corresponds to "daily," and \>30 corresponds to "multiple times per day."
:::

::: {#fig-signal_alluvial_plot}
```{r}
signal_alluvial_plot
```

Alluvial plot showing the combinations of signaling frequencies per month. Alluvium width equals the number of individiuals with the same frequencies for the three signals/states. Stratum height equals the count of individuals at a given frequency. Frequency key: Low includes "never" and "once a month or less," medium includes "more than once a month but less than once a week" and "more than once a week but not daily," high includes "daily," and "multiple times per day."
:::

::: {#fig-barplot_ConflictFreq}
```{r}
barplot_conflict
```

Stacked bar plot showing the frequency of conflict (per month) between child and primary caregiver as reported by primary caretaker. Color corresponds to sex of the child. Frequency key: 0 corresponds to "never," \<=1 corresponds to "once a month or less," 2-3 corresponds to "more than once a month but less than once a week," 4-29 corresponds to "more than once a week but not daily," 30 corresponds to "daily," and \>30 corresponds to "multiple times per day."
:::

::: {#fig-barplot_AlloparentingFreq}
```{r}
barplot_alloparenting
```

Stacked bar plot showing the frequency of a child's alloparenting effort (per month) as reported by primary caretaker. Color corresponds to sex of the child. Frequency key: 0 corresponds to "never," \<=1 corresponds to "once a month or less," 2-3 corresponds to "more than once a month but less than once a week," 4-29 corresponds to "more than once a week but not daily," 30 corresponds to "daily," and \>30 corresponds to "multiple times per day."
:::

## Signaling behavior is negatively associated with age

For every signaling measure except sadness (which also involves non-signaling functions), there was a statistically significant negative relationship between the age of the child and the frequency of signaling when controlling for the sex of the child, household demographics, the frequency of conflict with one's primary caretaker, caretaker perceptions of child susceptibility to illness, and an alloparenting effort by sex interaction. See @tbl-main_models for regression statistics and @fig-effect_age for effects plots.

Age had largest effect on crying (`r stats$mcryH$ChildAge$str`), with the model predicting crying frequency dropping to half of the age 5 value by around age 8 (visual inspection). The effect of age on temper tantrums (`r stats$mtantrumH$ChildAge$str`) was smaller with a more linear slope.

```{r}
#| label: tbl-main_models
#| tbl-cap: Regression tables for our main exploratory models.
r_vars <- c(
  ChildAge = "Age (years)",
  SexMale = "Sex (male)",
  OtherChildrenHH = "Other children",
  LogIncome = "Income (log)",
  number_adults = "Adults in home",
  PartnerStatusUnpartnered = "Partner status (single)",
  EducationLevelYears = "Parent ed. (years)",
  ConflictFreqN = "Conflict freq.",
  SignalFreq = "Signal freq.",
  SignalCost = "Signal cost",
  SignalFreqMax = "Max signal freq.",
  IllnessSusceptibilityMean = "Percieved illness susceptibility",
  SadFreqN = "Sadness freq.",
  CryFreqN = "Crying freq.",
  TantrumFreqN = "Tantrum freq.",
  AlloparentingFreqN = "Child allocare"
)
modelsummary(list(Sadness = msadH, Crying = mcryH, Tantrum = mtantrumH, SignalFreq = msfH, SignalCost = mscH, SignalFreqMax = mmsfH), gof_map = "nobs", coef_rename = r_vars, stars = c('*' = .05, '**' = .01, '***' = .001), coef_omit = "(Intercept)")
```

::: {#fig-effect_age}
```{r}
signaling_data_plot_age
```

The effect of age on specific signal frequencies, summed signal frequency, the frequency of a child's most common signal, and the summed cost of signaling behavior, all per month. Shaded areas are 95% CIs (only taking the uncertainty of fixed-effect parameters into account). Rug corresponds to actual values of child age with a small amount of jitter to reveal overlapped values. See @tbl-main_models for the corresponding regression table.
:::

## Signaling behavior is positively associated with conflict with primary caretakers

There were statistically significant positive relationships between the frequency of conflict between a child and their primary caretaker and all of our signaling measures when controlling for the sex of the child, the age of the child, household demographics, caretaker perceptions of child susceptibility to illness, and an alloparenting effort by sex interaction. See @tbl-main_models for regression statistics @fig-effect_conflict for effects plots.

::: {#fig-effect_conflict}
```{r}
signaling_data_plot_conflict
```

The effect of the frequency of conflict with one's primary caretaker on specific signal frequencies, summed signal frequency, the frequency of a child's most common signal, and the summed cost of signaling behavior, all per month. Shaded areas are 95% CIs (only taking the uncertainty of fixed-effect parameters into account). Rug corresponds to actual values of the frequency of conflict between child and primary caretaker with a small amount of jitter to reveal overlapped values. See @tbl-main_models for the corresponding regression table.
:::

## Signaling behavior is negatively associated with caregiver education

There were statistically significant negative relationships between years of caregiver education and all of our signaling measures except sadness (`r stats$msadH$EducationLevelYears$str`) when controlling for the sex of the child, the age of the child, household demographics, caretaker perceptions of child susceptibility to illness, and an alloparenting effort by sex interaction. See @tbl-main_models for regression statistics @fig-effect_edu for effects plots.

::: {#fig-effect_edu}
```{r}
signaling_data_plot_education
```

The effect of years of caregiver education on specific signal frequencies, summed signal frequency, the frequency of a child's most common signal, and the summed cost of signaling behavior, all per month. Shaded areas are 95% CIs (only taking the uncertainty of fixed-effect parameters into account). Rug corresponds to actual values of primary caretaker education (years) with a small amount of jitter to reveal overlapped values. See @tbl-main_models for the corresponding regression table.
:::

## Predictors of conflict between children and caretakers

The frequency of conflict between children and caretakers is negatively associated with child age (`r stats$mconflict$ChildAge$str`) and the number of adults who provide childcare in a household (`r stats$mconflict$AdultsChildcare$str`) when controlling for the sex of the child, household demographics, the number of adults who provide and do not contribute to childcare, an alloparenting effort by sex interaction, and the summed frequency of alloparenting from all other children in the household. See @tbl-conflict_dep_var for the corresponding regression table and @fig-effect_conflict_dep_var for effects plots.

There was also a negative relationship between conflict frequency and the number of older children, however, this was not statistically significant (`r stats$mconflict$OlderKids$str`). Similarly, the model predicted conflict to be more common between caretakers and males, but this also failed to reach statistical significance (`r stats$mconflict$SexMale$str`).

::: {#fig-effect_conflict_dep_var}
```{r}
signaling_data_plot_conflict_dep_var
```

The effect of A) child age and B) the number of adults in the household that contribute to childcare on the frequency of conflict between caretaker and child. Shaded areas are 95% CIs (only taking the uncertainty of fixed-effect parameters into account). Rug corresponds to actual values of A) child age and B) the number of adults in the household that contribute to childcare with a small amount of jitter to reveal overlapped values. See @tbl-conflict_dep_var for the corresponding regression table.
:::

```{r}
#| label: tbl-conflict_dep_var
#| tbl-cap: Regression table for the model examining the frequency of conflict between child and caretaker as the dependent variable.
r_vars <- c(
  ChildAge = "Age (years)",
  SexMale = "Sex (male)",
  OtherChildrenHH = "Other children",
  LogIncome = "Income (log)",
  number_adults = "Adults in home",
  PartnerStatusUnpartnered = "Partner status (single)",
  EducationLevelYears = "Parent ed. (years)",
  ConflictFreqN = "Conflict freq.",
  SignalFreq = "Signal freq.",
  SignalCost = "Signal cost",
  SignalFreqMax = "Max signal freq.",
  IllnessSusceptibilityMean = "Percieved illness susceptibility",
  SadFreqN = "Sadness freq.",
  CryFreqN = "Crying freq.",
  TantrumFreqN = "Tantrum freq.",
  AlloparentingFreqN = "Child allocare",
  TricepR = "Tricep skinfold (residuals)",
  SubscapR = "Subscapular skinfold (residuals)",
  BodyFat = "BodyFat (composite)",
  RelativeNeed3 = "Relative need (w/in household)",
  UnwantedTask = "Unwanted task",
  Punishment21 = "Punishment",
  Family21 = "Family conflict",
  DiscomfortPainInjuryIllness = "Discomfort/Pain/Injury/Illness",
  CaregiverResponse = "Caregiver Response",
  OlderKids = "Older children",
  YoungerKids = "Younger children",
  AdultsNoChildcare = "Adults that do not provide childcare",
  AdultsChildcare = "Adults that provide childcare",
  OtherChildAlloparentingFreqN = "Other household child alloparetning freq."
)
modelsummary(mconflict, gof_map = "nobs", coef_rename = r_vars, stars = c('*' = .05, '**' = .01, '***' = .001), coef_omit = "(Intercept)")
```

## Interaction between child sex and child alloparenting

For each signaling measure, there was also a significant interaction between how often a child cared for their siblings and the sex of the child when controlling for the age of the child, the sex of the child, household demographics, the frequency of conflict with one's primary caretaker, and caretaker perceptions of child susceptibility to illness. The slope of these relationships were positive for girls and flat or slightly negative for boys. In other words, our models predict that as children provide more allocare within their families, the frequency of signaling would increase with alloparenting effort for girls and remain at the same rate or decrease slightly with alloparenting effort for boys. See @tbl-main_models for the corresponding regression table and @fig-effect_allo_by_sex for effects plots.

::: {#fig-effect_allo_by_sex}
```{r}
signaling_data_plot_allo_by_sex
```

The effect of an alloparenting effort by sex interaction on specific signal frequencies, summed signal frequency, the frequency of a child's most common signal, and the summed cost of signaling behavior, all per month. Shaded areas are 95% CIs (only taking the uncertainty of fixed-effect parameters into account). Rug corresponds to actual values of child alloparenting frequency with a small amount of jitter to reveal overlapped values. See @tbl-conflict_dep_var for the corresponding regression table.
:::

## Positive association between child alloparenting effort and number of older children

There was a statistically significant negative relationship between the number of older children in a household and the frequency of child alloparenting effort `r stats$malloparenting$OlderKids$str` when controlling for the sex of the child, the number of younger children, household income (log), the number of older children, the number of adults who provide childcare within the household, and the summed frequency of alloparenting from all other children in the household. As our model could not handle high values of the summed frequency of other child alloparenting, we removed `r sum(d2$OtherChildAlloparentingFreqN > 121, na.rm = TRUE)` outliers (`r sum(d2$OtherChildAlloparentingFreqN == 360, na.rm = TRUE)` of which had a value of 360 times per month and `r sum(d2$OtherChildAlloparentingFreqN == 420, na.rm = TRUE)` of which had a value of 420 timers per month) despite expecting these values to not be outliers in the broader Utilian community. See @tbl-alloparenting_dep_var for the corresponding regression table and @fig-effect_alloparenting_dep_var_plot for effects plots.

::: {#fig-effect_alloparenting_dep_var_plot}
```{r}
# alloparenting_dep_var_plot
```

The effect of the number of older children in the household on the frequency of a child's alloparenting effort. Shaded areas are 95% CIs (only taking the uncertainty of fixed-effect parameters into account). Each dot is 1 child. Overlapping points are represented by the transparency of the point (lower transparency equals more values at a given coordinate). See @tbl-alloparenting_dep_var for the corresponding regression table.
:::

```{r}
#| label: tbl-alloparenting_dep_var
#| tbl-cap: Regression table for the model examining the frequency of conflict between child and caretaker as the dependent variable.
r_vars <- c(
  ChildAge = "Age (years)",
  SexMale = "Sex (male)",
  OtherChildrenHH = "Other children",
  LogIncome = "Income (log)",
  number_adults = "Adults in home",
  PartnerStatusUnpartnered = "Partner status (single)",
  EducationLevelYears = "Parent ed. (years)",
  ConflictFreqN = "Conflict freq.",
  SignalFreq = "Signal freq.",
  SignalCost = "Signal cost",
  SignalFreqMax = "Max signal freq.",
  IllnessSusceptibilityMean = "Percieved illness susceptibility",
  SadFreqN = "Sadness freq.",
  CryFreqN = "Crying freq.",
  TantrumFreqN = "Tantrum freq.",
  AlloparentingFreqN = "Child allocare",
  TricepR = "Tricep skinfold (residuals)",
  SubscapR = "Subscapular skinfold (residuals)",
  BodyFat = "BodyFat (composite)",
  RelativeNeed3 = "Relative need (w/in household)",
  UnwantedTask = "Unwanted task",
  Punishment21 = "Punishment",
  Family21 = "Family conflict",
  DiscomfortPainInjuryIllness = "Discomfort/Pain/Injury/Illness",
  CaregiverResponse = "Caregiver Response",
  OlderKids = "Older children",
  YoungerKids = "Younger children",
  AdultsNoChildcare = "Adults that do not provide childcare",
  AdultsChildcare = "Adults that provide childcare",
  OtherChildAlloparentingFreqN = "Other household child alloparetning freq."
)
modelsummary(malloparenting, gof_map = "nobs", coef_rename = r_vars, stars = c('*' = .05, '**' = .01, '***' = .001), coef_omit = "(Intercept)")
```

## Summed signal cost, crying frequency, and tantrum frequency are negatively associated with body fat

There were significant negative relationships between summed signal cost (`r stats$mBFc$BodyFat$str`) crying frequency (`r stats$mBFcry$BodyFat$str`), and tantrum frequency (`r stats$mBFtantrum$BodyFat$str`) and a scaled composite of subscapular and tricep skinfold thicknesses when controlling for the sex of the child, the age of the child, the frequency of conflict with one's primary caretaker, household demographics,caretaker perceptions of child susceptibility to illness, an alloparenting effort by sex interaction and a body fat by sex interaction. See @tbl-bodyfat_models for regression statistics and @fig-effect_bodyfat for effects plots. Subscapular and tricep skinfold thicknesses were also significant predictors of summed signal cost and tantrum frequency when included in the same models as the sole measure of body fat. See xxxxSI TABLESxxxx for regression tables.

Although not significant, our models predicted a negative slope for the relationship between body fat and summed signal cost, crying frequency, and tantrum frequency for girls but no relationship for boys.

Height Z-scores, weight Z-scores, BMI Z-scores, and body fat percentage residuals were not significant predictors of signaling behavior in models with the same control variables. See xxxxSI TABLESxxxx for regression tables.

```{r}
#| label: tbl-bodyfat_models
#| tbl-cap: Regression tables for the models examining the effect of body fat on signaling.
r_vars <- c(
  ChildAge = "Age (years)",
  SexMale = "Sex (male)",
  OtherChildrenHH = "Other children",
  LogIncome = "Income (log)",
  number_adults = "Adults in home",
  PartnerStatusUnpartnered = "Partner status (single)",
  EducationLevelYears = "Parent ed. (years)",
  ConflictFreqN = "Conflict freq.",
  SignalFreq = "Signal freq.",
  SignalCost = "Signal cost",
  SignalFreqMax = "Max signal freq.",
  IllnessSusceptibilityMean = "Percieved illness susceptibility",
  SadFreqN = "Sadness freq.",
  CryFreqN = "Crying freq.",
  TantrumFreqN = "Tantrum freq.",
  AlloparentingFreqN = "Child allocare",
  TricepR = "Tricep skinfold (residuals)",
  SubscapR = "Subscapular skinfold (residuals)",
  BodyFat = "BodyFat (composite)"
)
modelsummary(list(Sadness = mBFsad, Crying = mBFcry, Tantrum = mBFtantrum, SignalFreq = mBFf, SignalCost = mBFc, SignalFreqMax = mBFfm), gof_map = "nobs", coef_rename = r_vars, stars = c('*' = .05, '**' = .01, '***' = .001), coef_omit = "(Intercept)")
```

::: {#fig-effect_bodyfat}
```{r}
signaling_data_plot_bodyfat2
```

The effect of body fat (scaled composite measure of subscapular and tricep skinfold thicknesses) on A) the summed cost of signaling behavior, B) the frequency of crying, and C) the frequency of temper tantrums. Shaded areas are 95% CIs (only taking the uncertainty of fixed-effect parameters into account). Each dot is 1 child. Overlapping points are represented by the transparency of the point (lower transparency equals more values at a given coordinate). See @tbl-bodyfat_models for the corresponding regression table.
:::

## Crying frequency is positively associated with greater height and weight gain in a small longitudinal sample

Signaling data and longitudinal data on child height and weight (2023 and 2024) exist for `r sum(table(height_long_cry$frame$HeightMean_2024))` children. In this sample, there was a significant effect of a child age by crying frequency interaction on 2024 child height and weight whereby more crying was associated with more growth when controlling for 2023 height and weight.

See @tbl-longitudinal for the corresponding regression table.

# Check: SD (Observations) in the table

```{r}
#| label: tbl-longitudinal
#| tbl-cap: Regression tables for the models examining height and weight growth.
r_vars <- c(
  ChildAge = "Age (years)",
  SexMale = "Sex (male)",
  OtherChildrenHH = "Other children",
  LogIncome = "Income (log)",
  number_adults = "Adults in home",
  PartnerStatusUnpartnered = "Partner status (single)",
  EducationLevelYears = "Parent ed. (years)",
  ConflictFreqN = "Conflict freq.",
  SignalFreq = "Signal freq.",
  SignalCost = "Signal cost",
  SignalFreqMax = "Max signal freq.",
  IllnessSusceptibilityMean = "Percieved illness susceptibility",
  SadFreqN = "Sadness freq.",
  CryFreqN = "Crying freq.",
  TantrumFreqN = "Tantrum freq.",
  AlloparentingFreqN = "Child allocare",
  TricepR = "Tricep skinfold (residuals)",
  SubscapR = "Subscapular skinfold (residuals)",
  BodyFat = "BodyFat (composite)",
  RelativeNeed3 = "Relative need (w/in household)",
  HeightMean_2024 = "2024 height",
  HeightMean_2023 = "2023 height",
  WeightMean_KG_2024 = "2024 weight",
  WeightMean_KG_2023 = "2023 weight"
)

modelsummary(list(Weight2024 = weight_long_cry, Height2024 = height_long_cry), gof_map = "nobs", coef_rename = r_vars, stars = c('*' = .05, '**' = .01, '***' = .001), coef_omit = "(Intercept)")
```

::: {#fig-effect_longitudinal_plot}
```{r}
longitudinal_plot
```

The effect of a child crying frequency by age on child weight and height in 2024, controlling for the corresponding measure in 2023. Rug corresponds to actual values of crying frequency per month with a small amount of jitter to reveal overlapped values. See @tbl-longitudinal for the corresponding regression table.
:::

## Positive associations between child age and signaling frequency and child neediness

Due to the rarity of primary caretakers describing a child as "much less" (`r sum(na.omit(modeldf$RelativeNeed == "Much less"))` \[`r signif((sum(na.omit(modeldf$RelativeNeed == "Much less"))/ sum(!is.na(modeldf$RelativeNeed)) * 100), 2)`%\]) or "much more" (`r sum(na.omit(modeldf$RelativeNeed == "Much more"))` \[`r signif((sum(na.omit(modeldf$RelativeNeed == "Much more"))/ sum(!is.na(modeldf$RelativeNeed)) * 100), 2)`%\]) needy compared to their other children, we collapsed "less" and "much less" into one "less" factor and "more" and "much more" into one "more" factor. The end result was `r rel_need_table[1]` less needy children, `r rel_need_table[2]` children who were as needy as their siblings, and `r rel_need_table[3]` more needy children.

There were significant positive relationships between perceived relative neediness within the family and both the summed frequency of child signaling and child age when controlling for child sex, household demographics, the frequency of conflict between child and caretaker, and the frequency of a child providing allocare. There was also a positive relationship between summed signal cost and perceived relative neediness within the family when controlling for the same variables; however, this association was not significant. See @tbl-need_models for the corresponding regression table.

```{r}
#| label: tbl-need_models
#| tbl-cap: Regression tables for the models examining relative perceptions of child need within the family as a dependent variable.
r_vars <- c(
  ChildAge = "Age (years)",
  SexMale = "Sex (male)",
  OtherChildrenHH = "Other children",
  LogIncome = "Income (log)",
  number_adults = "Adults in home",
  PartnerStatusUnpartnered = "Partner status (single)",
  EducationLevelYears = "Parent ed. (years)",
  ConflictFreqN = "Conflict freq.",
  SignalFreq = "Signal freq.",
  SignalCost = "Signal cost",
  SignalFreqMax = "Max signal freq.",
  IllnessSusceptibilityMean = "Percieved illness susceptibility",
  SadFreqN = "Sadness freq.",
  CryFreqN = "Crying freq.",
  TantrumFreqN = "Tantrum freq.",
  AlloparentingFreqN = "Child allocare",
  TricepR = "Tricep skinfold (residuals)",
  SubscapR = "Subscapular skinfold (residuals)",
  BodyFat = "BodyFat (composite)",
  RelativeNeed3 = "Relative need (w/in household)"
)
modelsummary(list(SignalFreq = mN1, SignalCost = mN2), vcov = vcovCL, type = "HC0", cluster = ~householdID, gof_map = "nobs", coef_rename = r_vars, stars = c('*' = .05, '**' = .01, '***' = .001), coef_omit = "(Intercept)")
```

# FIX: error ribbons cross 1 and 0

::: {#fig-effect_relativeneed}
```{r, fig.width= 10}
signaling_plot_need
```

The likelihood of child being more needy, as needy, or less needy than a caretaker's other children. A: Effect of child age on relative neediness B: Effect of summed signaling frequency on relative neediness.
:::

## Caregiver responses to last signal of need

There were `r response_overlap[2,2]` instances in which a caregiver's response involved acts which were both negative and positive from the child's perspective. These were excluded from analyses due to their rarity.

There were no significant relationships between, child age, child sex, the number of children in the household, and log family income and the valence of parental reponses (positive, neutral, or negative) from the child's perspective. However, primary caretakers were significantly more likely to respond positively towards a child when they attributed the cause of signaling to discomfort, pain, injury, or illness (stat), and significantly less likely to respond positively when they attributed the cause of signaling to child punishment (stat). See @fig-effect_last_signal_response and @tbl-last_signal_model.

```{r}
#| label: tbl-last_signal_model
#| tbl-cap: Regression table for the model examining responses to the last time a child was sad or cried.
r_vars <- c(
  ChildAge = "Age (years)",
  SexMale = "Sex (male)",
  OtherChildrenHH = "Other children",
  LogIncome = "Income (log)",
  number_adults = "Adults in home",
  PartnerStatusUnpartnered = "Partner status (single)",
  EducationLevelYears = "Parent ed. (years)",
  ConflictFreqN = "Conflict freq.",
  SignalFreq = "Signal freq.",
  SignalCost = "Signal cost",
  SignalFreqMax = "Max signal freq.",
  IllnessSusceptibilityMean = "Percieved illness susceptibility",
  SadFreqN = "Sadness freq.",
  CryFreqN = "Crying freq.",
  TantrumFreqN = "Tantrum freq.",
  AlloparentingFreqN = "Child allocare",
  TricepR = "Tricep skinfold (residuals)",
  SubscapR = "Subscapular skinfold (residuals)",
  BodyFat = "BodyFat (composite)",
  RelativeNeed3 = "Relative need (w/in household)",
  UnwantedTask = "Unwanted task",
  Punishment21 = "Punishment",
  Family21 = "Family conflict",
  DiscomfortPainInjuryIllness = "Discomfort/Pain/Injury/Illness",
  CaregiverResponse = "Caregiver Response"
)
modelsummary(mpr, vcov = vcovCL, type = "HC0", cluster = ~householdID, gof_map = "nobs", coef_rename = r_vars, stars = c('*' = .05, '**' = .01, '***' = .001), coef_omit = "(Intercept)")
```

::: {#fig-effect_last_signal_response}
```{r, fig.width= 8}
signaling_plot_last_signal_response
```

The likelihood of positive, neutral, and negative responses to the last time a child became sad or cried. A: Effect of punishment being implicated in the cause (0 = no punishment, 1 = punishment.) B: Effect of discomfort, pain, injury, or illness being implicated in the cause (0 = discomfort, pain, injury, or illness, 1 = discomfort, pain, injury, or illness.) Legend key: -1 = negative response, 0 = neutral response, 1 = positive response (all from child's perspective).
:::

## Sociodemographics

We found no significant associations between sex and child signaling behavior outside of the previously mentioned interactions, including when sex was included in models as interaction with age. See table SI xxxxxxx for regression tables which included the sex by age interaction. Monthly household income also was not a significant predictor of signaling behavior in any model nor was population when using language, immigration status, and neighborhood as proxies for it. See table SI xxxxxxx.

# Discussion

*Negative association between age and child signaling (all signaling measures)*

As seen in Western cultures [@zeifman2013; @potegal2003; @koch2003] (check for others), the frequencies of crying and temper tantrums decreased with age, something which is likely due to numerous reasons. One possibility is that older children have greater ability to satisfy one's desires and behave in ways which match cultural expectations of teenagers and adults. Another is that since one's reproductive value increases as one survives periods of high mortality risk in childhood, the intensity of conflicts of interest with parents decreases since these children are likely to reach adulthood and require less investment to reach adulthood than younger children, current or future [@daly1995]. For this reason, less costly signals may be more likely to work to increase investment, all else being equal\[\[4\]\]. That said, all else may often not be equal if investment in older children is costlier, something which may be reflected in caretakers seeing older children as relatively needy. The decrease in crying and tantrum frequencies with age may also result from the potential for greater reputational costs, something which may also increase the credibility of such signals among older children as previously mentioned. ~~It is also possible that parents see older children less than younger children. However, the negative association between need with age observed in other societies and the fact that all children in this study were of school age suggest that this is unlikely to be the sole reason responsible for this.~~

There was no significant association between the frequency of child sadness and age. Part of this likely results from the fact that, unlike crying, sadness is thought to have important non-signaling functions including drawing one’s attention to a problem, motivating deeper analysis of one’s situation, and causing withdrawal from situations with low fitness prospects [@nesse2019; @nettle2009; @allen2003; @hagen2011], from which, others can still obtain cue-based information about one's emotional state, well-being, and desires. By asking about sadness as a whole, our measure was unable to disentangle information caretakers gleaned from cues from information that came from signals. For this reason, a change in signals that arise from sadness outside of crying (e.g., facial sadness and some, but all, forms of withdrawal) may have been harder to detect with our sample size even if young children expressed facial sadness more often. It is also possible that there is no effect of age on sadness frequency among the age ranges examined it.

-   add in depression over the lifecourse?

Negative association with signal cost too

*Positive association between conflict with parent and child signaling (all signaling measures)*

Consistent with signaling models which emphasize the role of conflicts of interest in favoring the use of costlier signals, there were positive associations between caretakers' perceptions of conflict frequency and 1) the summed costliness of children's signaling, 2) the frequency of child signals of need, and 3) the frequency of children's most common need signaling strategy. Although conflicts of interest do not always lead to behavioral conflict [@altmann1980; @bateson1994], it is likely many, if not most, conflicts are ultimately caused by conflicts of interest, making it a useful proxy for the underlying conflicts of interest that may increase the benefits of deception and the costs of being deceived\[\[5\]\]. If this is the case, the association between signaling and conflict may represent children's attempts at credibly signaling their value of additional investment or the costliness of punishment [@syme2019] to parents who would have good reason to doubt the trustworthiness of simple verbal requests (or the trustworthiness of facial sadness and crying for those resorting to temper tantrums).

*Sex and focal child alloparenting effort interact (positive for females, negative/flat for males)*

We found evidence of an interaction between how often a child cared for their siblings and the sex of the child in our models examining signal frequencies and cost, whereby there was a positive relationship between the frequency and summed costliness of child signals of need and alloparenting effort for girls but no relationship or a slightly negative one for boys. In many societies, girls are often expected to do more alloparenting within the household and often end up doing so [@helfrecht2016 @weisner1977; @barry1971; @turke1988]. While girls in this sample were also reported to give more allocare, the difference was not significant. However, Utilian mothers reported in focus groups (confirm) that girls tend to provide more help for their siblings than boys. If this difference was the outcome of caretakers pressuring girls to do more or resulted in the expectation of more alloparenting, the increase in signaling with alloparenting effort seen among girls may be the result of a strategy to signal the costliness of their efforts in order to invest more in romantic relationships and status seeking behaviors outside the household.

Evidence for conflicts similar to the one hypothesized here has been reported by multiple researchers working with the Chuukese. Lowe found that both male and female children reported conflict with their families over their desires to invest financial resources in their own status enhancing opportunities rather than in the household [@lowe2003]. In an analysis of conflict among Chuukese-Americans, Syme and Hagen found that severe conflicts were more likely with greater numbers of siblings and when conflicts involved a child spending time away from home or evading family responsibilities [@syme2023a], findings which would be consistent with parents pressuring older children into investing more in the household among other explanations. In response to such conflicts, participants reported using a wide range strategies as children including, deception, withholding cooperation, running away, and seeking the assistance of others. Through these means, participants were often able to resolve conflicts in their favor despite their lower status in the household and community.

*Alloparenting effort model* Not sure on interpretation here yet

*Body fat discussion*

Children with less body fat cried more, threw more temper tantrums, and had higher summed signal costs. In other words, needier children signaled more often. Although the body fat by sex interaction was only marginally significant, our models predict a negative relationship between body fat and signaling for girls and no relationship for boys. Female reproduction requires a significant caloric invest in children, both in the womb and when breastfeeding [@jasienska2009], costs which males do not have to pay. For this reason, low body fat, which can prevent ovulation in extreme cases [@ellison2003], is likely a be better predictor of future reproductive success in adolescent (and older) girls than boys, potentially making signaling for increased investment a more viable strategy in instances when an individual has low body fat.

*Positive associations between child age and signaling frequency and child neediness*

There was a positive association between the summed frequency of child signaling and caretaker perceptions of child neediness within the family, something which would be consistent with sadness, crying, and temper tantrums being signals (or involving signals in the case of crying) if parents took this to mean how much they value additional investment. Caretakers also perceived older children as needier, despite having more conflicts with younger children. One reason for this could be that caretakers weighted monetary investment highly when evaluating child need.

Signaling behavior did not predict (try more models?) differences in relative caregiver investment between children within households. However, crying frequency was associated with both height and weight gain for older children. If older children face greater reputation costs from crying, crying may be a relatively credible credible signal compared to crying among younger children, with it being a less viable strategy to exaggerate one's value of investment or the costliness of adversity.

*Parental responses to last signal of need*

Parents were more likely to respond positively and less likely to respond negatively to the last time a child signaled when the cause involved discomfort, pain, injury, or illness than when it did not. One potential causal explanation for this is that parents may often have information for sources outside of child signaling to corroborate a child's need (e.g., a fever, a diagnosis, or witness a child hurt themselves) so that there is less private information about a child's current state. Simultaneously, conflicts of interest may be less significant in these times due to either 1) both sides benefiting similarly from their child returning to normal (e.g., with an illness) or 2) the cause involving something which is minor and would quickly resolve itself (e.g. a skinned knee). It may also be that many of the typical positive responses to pain and illness are not very costly to caretakers (e.g., providing comfort), thereby increasing their frequency.

When the cause involved punishment, however, parents were more likely to respond negatively and less likely to respond positively. In these circumstances, children may be signaling about the severity of a punishment [@syme2019]. However, parents may be reluctant to reduce punishments if the behavior that warranted the punishment was repeated or severe. They may also react with scolding or further punishment. That said, our model predicts positive responses in `r signif(updated_out$estimate[3] * 100, 2)`% of cases (when all other predictors are held at their mean values), suggesting that children are successful in negotiating better outcomes at least some of the time.

Although not a significant relationship, the negative effect of the family conflict being a reported cause of signaling on the likelihood of a positive response may reflect tradeoffs within the family, whereby investment in one child cannot go to another. Caretakers may have also responded more negatively when the cause involved family conflict, if it involved a child imposing costs on other family members.

*In most models, there is a negative association between education and child signaling?*

We found a negative association between caretakers’ education level (years in school) and many of our child signaling measures despite our models controlling for household income and partner status. One possible reason for this may be that more educated caretakers see their children less during the day due to greater ability to afford childcare or due to them having jobs that are less compatible with child care (test model of social support predicted by education). Unfortunately, we do not have data on how much time caretakers spent with children to test this, and it seems equally likely that less educated parents may see their kids less often if they need to work more hours to support their families. Another possibility is that lower levels of education are associated with greater levels of adversity for both parents and children, something which appears to be the case for other measures of SES [@pepper2017] and would increase the benefits of signaling need for children and the difficulty of parents to respond. It is also possible that education influences parenting norms in ways that cause differences in child signaling behavior.

*Lack of evidence for demographic effects*

Outside of the sex by alloparenting effort interaction, we found little evidence of sex differences in child signaling behavior. As adults, females cry more than males, a finding that has been reported across cultures [@vanhemert2011; @becht2001]. One reason why our study may have failed to find significant effects of sex is that sex differences in the causes and reputational effects of signaling need are small or nonexistent until adolescence and young adulthood, with our small sample size for this age range preventing us from detecting a real effect should it exist. It is also possible that the sex difference seen in rates of crying in adults may reflect the lesser ability of females to bargain for better treatment through aggression or the threat of it when faced with adversity [@hagen2016]. Such a sex difference may not exist among preadolescencents due to a lack of difference in strength or athletic ability beofre puberty [@handelsman2018] or due to their relative lack of formidability compared to adults.

Monthly income was also not found to be a significant predictor of signaling behavior. Many models designed to explain signals of need (e.g., the begging of young birds) predict that offspring who are worse off or are receiving less investment have more to gain from signaling than those with more [@mock2011; @godfray1991; @caro2016a; @godfray2000]. However, it is possible that this can be true and household level resources have little influence on child signaling frequencies between families. As signaling behavior is costly to child, caretakers, and siblings, children may gain direct and indirect fitness benefits from limiting their signaling behavior in times of resource stress (i.e., when the chances of gaining investment or the size of investment are small). Similarly, children in high resource families may also be motivated to signal at high frequencies due to the potential for larger pieces of investment. We know of no empirical explorations of this in humans (check more and confirm). However, multiple game theoretical models examining parent-offspring conflict have found that parental resources have a modest impact on the evolutionary stable value for offspring demand for resources [@bossan2013; @mock1997 [confirm]]. ~~Although counterintuitive, such an outcome fits the fact that fitness is relative and that natural selection should be expected to have favored those who maximized their own inclusive fitness, whether one found themselves in a great situation or a dire one.~~ Compared to overall effort in eliciting need, resource availability may have a larger impact on how children signal for more investment based on the finding that birds in stable and unusually favorable environments are more likely to compete with siblings through signals thought to reveal information about need, while those in relatively unpredictable and unusually poor environments have been found to compete more through signals thought to reveal information about quality [@caro2016a].

Despite population level differences in wealth, job certainty, discrimination, and living conditions [@davis2014; @garcia2017; @garcia2018; @stonich2000], we generally did not find an effect of population on child signaling behavior. Although we sought data from both English-speaking participants born on the island and Spanish-speaking immigrants from the mainland, our smaller sample of English-speaking Utilians may have limited our ability to detect real differences between the populations. It is possible that differences in wealth and status between the groups translate to differences in child signaling due to children facing different types adversity or weighting different types of investment differently. It is also likely that the two groups differ in how caretakers could respond to child signaling, and caretakers between the two populations may also differ in their cultural models of parenting. However, as previously mentioned, it is conceivable that household level differences have little effect on child signaling behavior.

## Limitations

The main limitation of this study is that it is completely exploratory, something which results in a high risk of our models overfitting the data and indicating associations which might not exist in the real world. Our study was also predominantly cross-sectional, further limiting our ability to make causal inferences.

Our use of convenience and snowball sampling is also a limitation as it resulted in a non-probability sample which may not be representative of the broader population. This could happen if there was selection bias so that those who participated differed from the rest of the population in how they interacted with their children or how they assessed, remembered, and reported child behavior.

Outside of child anthropometric measurements, all data come from caregiver reports, and therefore, reflect the biases of caretakers, whatever they may be. Due to the fact the data used here were a part of survey which covered many topics and often multiple children, we also faced a tradeoff between survey time and question depth. Asking adult participants about their relative investment in children in their household was likely not as effective as more in depth, questions focused on specific questions. Framing the question this way may have also increased the likelihood of social desirability bias [@grimm2010], which could lead adult participants to be be more likely to report equal investment across children. Our questions about the last time the child signaled may have also been more effective if they focused on specific signals due to our methods combining crying, which is unambiguous as a signal, with sadness, which has both signaling and non-signaling functions.

**Footnotes**

[\[1\]](#_ftnref1) We expect “signals of need” to actually be signals of resource valuation. However, we use “signal of need” here to distinguish signals that involve attempts at eliciting support through revealing private information about one’s vulnerabilities from the more commonly studied signals which reveal private information about one’s positive qualities (i.e., those related to partner choice and intrasexual competition).

[\[2\]](#_ftnref2) Our use of “credible” signaling is largely synonymous with the more common term, honest signaling. However, we prefer the term credible signals since signals do not need to be completely honest for natural selection to have favored individuals who paid attention to them.

[\[3\]](#_ftnref3) Researchers often distinguish between sadness and low mood based on the length of negative affect. We remain agnostic on how different sadness and low mood are from one another but use the terms researchers used when describing their research.

[\[4\]](#_ftnref4) All else is often not equal when it comes to the costliness of investment as older children may often cost more (e.g., the costs education and food are often higher among older children). However, parents would still have to pay similar costs to younger children as they grew, something which natural selection has likely acted on when it comes to our adaptations involved with differential child investment.

[\[5\]](#_ftnref5) Even in instances where conflicts begin due to mistakes, the initial conflict can lead to increased conflicts of interest. This may be seen with findings that individuals who accidentally harm another often grow to dislike them (cite), something which has been suggested to be adaptive to the accidental perpetrator as the unintended victim is likely to value than they would have before, potentially to the point of wishing them harm.

[\[6\]](#_ftnref6) Despite the risk, Shostak notes there were no known recent incidents of harm to women who ran away to spend the night in the Bush [@shostak1981].

# Supplementary Information

List of options for research assistant to classify participant responses to the question of how they responded when their child was last sad or cried.

Positive: Give resources (e.g., toys, food, money)\
Positive: Promise resources\
Positive: Give attention/affection\
Positive: Promise attention/affection\
Positive: Give opportunities (e.g., loosen restrictions, new privileges, time with friends, fun activities)\
Positive: Promise opportunities\
Positive: Remove responsibilities (e.g., fewer chores)\
Positive: Promise removal of responsibilities\
Positive: Take child's side in conflict (e.g., taking child's side against sibling, intervening against bully)\
Positive: Give advice\
Positive: Other \[Text Box\]\
Neutral: Distract\
Neutral: Do nothing\
Neutral: Explain (e.g., go over why the child cannot have or do something)\
Neutral: Other \[Text Box\]\
Negative: Threaten removal of resources\
Negative: Remove resources (e.g., toys, food, money)\
Negative: Threaten withholding of attention/affection\
Negative: Withhold attention/affection (e.g., silent treatment, ignoring)\
Negative: Threaten loss of opportunities (e.g., freedom to move around the island, beach time, timeout, time with friends)\
Negative: Loss of opportunities\
Negative: Threaten more responsibilities (e.g., chores, alloparenting, homework)\
Negative: Add responsibilities\
Negative: Support adversary (e.g., taking sibling's side)\
Negative: Threaten physical punishment\
Negative: Punish physically\
Negative: Punish verbally (includes yelling)\
Negative: Downplay adversity\
Negative: Other \[Text Box\]\
NA (Can't remember, refuse to answer, etc)\
Other: Other \[Text Box\]
